---
title: "Gov 50 Final Project"
author: "Joshua Parker"
description: "Analyzing the Impact of Base MSRP and Driving Range on Electric Vehicle Adoption"
output:
  distill::distill_article:
    self_contained: false
---

## Project Thoughts

I am interested in exploring data related to the impact of the base Manufacturer's Suggested Retail Price (MSRP) and driving range of electric vehicles (EVs) on the adoption of EVs in cities. Understanding the relationships between the MSRP and driving range of EVs and consumer behavior is critical in helping manufacturers and policymakers develop strategies to make EVs more appealing and accessible.

## Research Proposal

Does the base MSRP and driving range of EVs have a causal effect on the adoption of EVs in cities?

I hypothesize that cities with lower base MSRP and higher driving range EVs are more likely to experience increased adoption of EVs. This is rooted in the belief that affordability and a strong battery range are attractive to the common customer. In contrast, cities with higher base MSRP and lower driving range EVs may lack the economic accessibility to practical EVs, and thus, witness lower rates of EV adoption.

All research will be conducted using the publicly available Electric Vehicle Population data set from the Washington State Department of Licensing (DOL) and US Zip Codes Database from the US Census Bureau.

The explanatory variables of interest are base MSRP (base_msrp) and driving range (electric_range). Base MSRP is measured as the lowest MSRP for any trim level of the model in question. Driving range is measured as the distance the vehicle can travel purely on its electric charge.

The outcome variable of interest is EV adoption in cities. This variable is not present in the raw data set, but will be measured as the count of registered EVs in a given city.

If the analysis yields a negative and statistically significant coefficient for the base MSRP variable and a positive and statistically significant coefficient for the driving range variable, it would provide support for the hypothesis. This would suggest that, on average, EVs with lower base MSRPs and longer electric ranges are associated with a higher count of EVs in cities. Such a pattern would align with the idea that addressing high costs and range anxiety through longer electric ranges positively influences consumer adoption decisions.

Conversely, if the coefficient for base MSRP is not statistically significant or positive, or the coefficient for driving range is not statistically significant or negative, it would suggest that there is no clear association between base MSRP, electric range and electric vehicle adoption. This could imply that other factors, such as vehicle design and charging infrastructure, play a more prominent role in consumer adoption trends.

## Data Wrangling

```{r}
# load packages
library(tidyverse)
library(dplyr)
library(ggplot2)
library(maps)

# read csv files
EV <- read_csv("EV_data.csv")
zipcode <- read_csv("zipcode_data.csv")

# drop rows where base MSRP or driving range equal zero
EV <- EV %>%
  filter(base_msrp != 0,
         electric_range != 0,
         state == "WA")

# drop columns that are not latitude or longitude
zipcode <- zipcode %>%
  select(zip, lat, lng) %>%
  rename(zip_code = zip) %>%
  rename(long = lng) %>%
  mutate(zip_code = as.numeric(zip_code))

# merge EV and zipcode datasets
EV <- left_join(EV, zipcode, by = "zip_code")

# data grouped by city
cities <- EV %>%
  group_by(city) %>%
  summarize(city_EV_count = n(),
            average_base_msrp = mean(base_msrp),
            average_electric_range = mean(electric_range),
            lat = mean(lat),
            long = mean(long)) %>%
  arrange(desc(city_EV_count))

# map of washington state
washington_map <- map_data("state", region = "washington")
```

## Data Visualizations

```{r}
# graph 1: city EV count density map
ggplot() +
  geom_polygon(washington_map, 
              mapping = aes(x = long,
                            y = lat),
              color = "black",
              fill = "white") +
  geom_point(cities,
             mapping = aes(x = long,
                           y = lat,
                           size = city_EV_count),
             alpha = 0.3,
             color = "blue") +
  scale_size(range = c(1,10)) +
  labs(title = "Graph 1: City EV Count Density Map",
       x = "Longitude",
       y = "Latitude",
       size = "EV Count") +
  theme_minimal()
```

```{r}
# graph 2: city EV count histogram
ggplot(cities, aes(x = city_EV_count)) +
  geom_histogram(bins = 40,
                 fill = "blue") +
  labs(title = "Graph 2: City EV Count Histogram",
       x = "City EV Count",
       y = "Count") +
  theme_minimal()
```

```{r}
# graph 3: base MSRP histogram
ggplot(EV, aes(x = base_msrp)) +
  geom_histogram(fill = "red3",
                 bins = 40) +
  xlim(c(0, 200000)) +
  labs(title = "Graph 3: Base MSRP Histogram",
       x = "Base MSRP ($)",
       y = "Count") +
  theme_minimal()
```

```{r}
# graph 4: driving range histogram
ggplot(EV, aes(x = electric_range)) +
  geom_histogram(fill = "darkgreen",
                 bins = 40) +
  labs(title = "Graph 4: Driving Range Histogram",
       x = "Driving Range (mi.)",
       y = "Count") +
  theme_minimal()
```

```{r}
# graph 5: average base MSRP vs city EV count scatter plot
ggplot(cities, aes(x = average_base_msrp,
                   y = city_EV_count)) +
  xlim(30000, 110000) +
  ylim(0, 200) +
  geom_point(color = "red3",
             alpha = 0.3,
             size = 2) +
  geom_smooth(method = "lm",
              se = FALSE) +
  labs(title = "Graph 5: Average Base MSRP vs City EV Count",
       x = "Average Base MSRP ($)",
       y = "City EV Count") +
  theme_minimal()
```

```{r}
# graph 6: average driving range vs city EV count scatter plot
ggplot(cities, aes(x = average_electric_range,
                   y = city_EV_count)) +
  geom_point(color = "darkgreen",
             alpha = 0.3,
             size = 2) +
  geom_smooth(method = "lm",
              se = FALSE) +
  labs(title = "Graph 6: Average Driving Range vs City EV Count",
       x = "Average Driving Range (mi.)",
       y = "City EV Count") +
  theme_minimal()
```

```{r}
# table 1: correlation table
cor_table <- tibble(base_msrp_cor = cor(cities$average_base_msrp,
                                        cities$city_EV_count),
                    electric_range_cor = cor(cities$average_electric_range,
                                             cities$city_EV_count))

knitr::kable(cor_table,
             digits = 4,
             col.names = c("Base MSRP", "Driving Range"),
             caption = "Correlation Coefficients")
```

## Analysis

The research design is cross-sectional, utilizing observational data from November 15, 2023. This offers a snapshot at the relationships between base MSRP, driving range, and EV adoption in Washington cities.

#### Variable Plots (Graphs 1-4)

The majority of cities have less than 100 EVs, with most high-density cities in the Seattle metropolitan area. Cities with high EV-counts tend to be close to other high EV-count cities, forming clusters. This may be a result of EVs being more convenient when nearby cities also have a high concentration of EVs and EV-friendly infrastructure. The right-skew in the city EV count histogram reflects a concentration of EVs in a few cities, likely due to higher income levels or more charging stations (Graphs 1-2).

Base MSRP is relatively similar across EVs, ranging from \$25k to \$75k, which suggests a competitive pricing strategy to align with gasoline cars. The outlier at \$175k might represent luxury EVs targeted at wealthier individuals (Graph 3).

The sporadic nature of EV driving ranges and the clusering around \~20, \~90, and \~210 miles might reflect different tiers of EV technology and consumer preferences. For example, EVs might use a few types of batteries that each have a specific range. Moreover, car body style (ie. coupe, sedan, SUV) might play a role in the congregation of driving ranges (Graph 4).

#### Regressions (Graphs 5-6, Table 1)

The hypothesis of this experiment is that cities with lower base MSRP and higher driving range EVs have higher EV counts.

The correlation coefficient of base MSRP and city EV count is -0.0199; on average, for every \$1 increase in average base MSRP, there is a 0.0199 decrease in the number of EVs in a city. The negative coefficient supports the hypothesis that base MSRP and city EV count are negatively correlated. However, the low coefficient value is not statistically or practically significant, suggesting some price sensitivity among consumers but not as a predominant factor (Graph 5, Table 1).

The correlation coefficient of driving range and city EV count is -0.0486; on average, for every 1 mile increase in driving range, there is a 0.0486 decrease in the number of EVs in a city. The negative coefficient reveals a counterintuitive trend that disproves the hypothesis, implying that a higher driving range results is a lower city EV count, albiet the effect size relatively small. Similar to base MSRP, this suggests that longer ranges are not necessarily a decisive factor for consumers (Graph 6, Table 1).

The cross-sectional design of this experiment captures correlations at a certain point in time. Therefore, we should not interpret these results causally.

## Conclusion

The results of this experiment provide mixed support for the hypothesis. While lower base MSRP and higher driving range are believed to encourage EV adoption, the observed correlations suggest that these factors play a less significant role than expected. Moreover, it emphasizes that government policies to slash EV purchase prices or manufacturer initiatives to increase driving ranges may not significantly increase EV adoption by themselves. A more multi-faceted approach is necessary.

The cross-sectional design and sole focus on Washington makes it difficult to interpret these results causally and extrapolate them to the general population. Factors such as regional policies, charging infrastructure, or socioeconomic standings have a strong influence on EV adoption patterns and may have significantly confounded the results.

With more time and money, the analysis could be improved through an experimental study on a broader geographic area. In this improved study, I would also incorporate multivariable regressions to control for potential confounding factors. This would allow us to interpret conclusions causally and generalize the results to the public.
